A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2020-04-17 paper/paper-revision.Rnw
A few introduction tweaks, add SWGTREAD citation

\begin{document}
\maketitle
\begin{abstract}
Quantitative assessment of class characteristic frequency is a difficult problem, in part because there is currently no way to collect and efficiently process the quantity of data that would be required for such assessments without prohibitive amounts of human labor. In this paper, we discuss a computational method which complements a passive, unmanned scanning device that is currently under development. The automatic class characteristic recognition algorithm described in this paper has been developed with photographs of outsole images, and identifies high-level \mt{geometric} features such as circles, stars, and quadrilaterals. These features are designed to mimic high-level features used by examiners to facilitate automatic processing of features without resorting to black-box feature sets commonly used in neural networks. The algorithm performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant step toward enabling the collection and application of population-level class characteristic data to support quantitative assessments of the value of class-characteristic based similarity assessments.
\end{abstract}
%
% \begin{keyword}
% Footwear; class characteristics; \mt{random match probability}; computer vision; neural networks
% \end{keyword}

@@

Collection and use of footwear evidence varies heavily across jurisdictions within the United States, and, though it is more common for such analyses to be used in Europe, the U.K., and Israel, many of the same fundamental issues remain. It is often difficult to recover footwear evidence with sufficient detail to identify individualizing characteristics, either because of degradation or because the medium did not record the individualizing characteristics in sufficient detail to examine. It is also common for prints to be only partially recorded or recovered, which further compounds the difficulty of performing these types of analysis. As a result, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. This tendency is also reflected in the SWGTREAD range of conclusions scale, which has two separate categories to indicate degrees of class characteristic based association out of a range of 7 conclusions.\fix{add citation} Currently, the guidelines also caution against any statistical assessment because accurate data and models to support statistical calculations do not exist. \mt{However,} if it were possible to collect a sufficiently large quantity of footwear impression data from a local population, it would be helpful to be able to calculate the frequency of shoes \mt{in the population that have similar class characteristics to a given shoe or print in order to characterize the determinative value of that} particular piece of evidence. Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns on the class characteristic level. %\fix{[This is an important point. Should I make a bigger deal out of it throughout the paper?]}% if it fits in, that's a good idea.

Statistically, the coincidental (or random) match probability is estimated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print. The primary obstacle to coincidental match probability calculation, then, is one of data collection and annotation.
The (limited) number of population class-characteristic studies \citep{parent_sandy_significance_nodate,benedict_geographical_2014} demonstrate the effort required to collect, process, and analyze even a moderate amount of data manually. However, technological innovations over the past decade have automated previously manual tasks such as license plate reading, tracking individuals on video feeds, and other complex tasks. In these comparable situations, two separate but related technological innovations were required: inexpensive and automatic data collection (e.g. via CCTV or video cameras) hardware, and automatic, accurate data processing software.
We are in the process of developing a low-profile scanner which, when embedded in a pedestrian area, passively collects images of shoe treads (a prototype is shown in \autoref{fig:prototype-scanner}). In parallel, however, it is necessary to develop software which can process the data from a similarly oriented scanner to generate features which are used to make class characteristic comparisons. Through the combination of hardware and software, we hope to enable the collection of population-level footwear data which will facilitate quantitative assessment of class characteristic frequency to support examiner evaluation.

@@

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner suggest that cameras are more robust than impression-based systems (e.g.  EverOS) for use in outdoor settings. As a result, in this paper describing an initial automated software system for class characteristic recognition, we use outsole images (such as those found on online retail sites), rather than impression-based training data. While we recognize that images from the scanner are likely to show some degradation compared to images found from online retailers, the fundamental features of the underlying machine learning models are similar. In this paper, we present a proof-of-concept model designed to identify class characteristics probabilistically, using labels similar to those used by examiners. We discuss the strengths and limitations of both the classification scheme and the neural-network based model, with important implications for future attempts to develop automatic high-level class characteristic recognition systems.

>>>>

\begin{document}
\maketitle
\begin{abstract}
Quantitative assessment of class characteristic frequency is a difficult problem, in part because there is currently no way to collect and efficiently process the quantity of data that would be required for such assessments without prohibitive amounts of human labor. In this paper, we discuss a computational method which complements a passive, unmanned scanning device that is currently under development. The automatic class characteristic recognition algorithm described in this paper has been developed with photographs of outsole images, and identifies high-level geometric features such as circles, stars, and quadrilaterals. These features are designed to mimic high-level features used by examiners to facilitate automatic processing of features without resorting to black-box feature sets commonly used in neural networks. The algorithm performs well on images with unambiguous shapes and moderate color contrast, though some systematic errors arise from the physical similarities of geometric shapes. Additional improvements may be realized by modifying the labeling scheme to reduce feature overlap and by preprocessing the images to improve contrast. This new approach represents a significant step toward enabling the collection and application of population-level class characteristic data to support quantitative assessments of the value of class-characteristic based similarity assessments.
\end{abstract}
%
% \begin{keyword}
% Footwear; class characteristics; \mt{random match probability}; computer vision; neural networks
% \end{keyword}

@@

Collection and use of footwear evidence varies heavily across jurisdictions within the United States, and, though it is more common for such analyses to be used in Europe, the U.K., and Israel, many of the same fundamental issues remain. It is often difficult to recover footwear evidence with sufficient detail to identify individualizing characteristics, either because of degradation or because the medium did not record the individualizing characteristics in sufficient detail to examine. It is also common for prints to be only partially recorded or recovered, which further compounds the difficulty of performing these types of analysis. As a result, the majority of footwear examinations in the United States reach conclusions regarding the correspondence of class characteristics rather than of individualizing characteristics \citep{smith_2011}. This tendency is also reflected in the SWGTREAD range of conclusions scale, which has two separate categories to indicate degrees of class characteristic based association out of a range of 7 conclusions\citep{SWGTREAD2013}. Currently, the guidelines also caution against any statistical assessment because accurate data and models to support statistical calculations do not exist. However, if it were possible to collect a sufficiently large quantity of footwear impression data from a local population, it would be helpful to be able to calculate the frequency of shoes in the population that have similar class characteristics to a given shoe or print in order to characterize the determinative value of that particular piece of evidence. Note that this question is distinct from make and model determination since many shoes of different makes and models can still have similar tread patterns on the class characteristic level.

Statistically, the coincidental (or random) match probability is estimated by sampling a large number of shoes from a given population and calculating the observed proportion of shoes in the sample whose tread patterns are similar to the given print. The primary obstacle to coincidental match probability calculation, then, is one of data collection and annotation. The (limited) number of population class-characteristic studies \citep{parent_sandy_significance_nodate,benedict_geographical_2014} demonstrate the effort required to collect, process, and analyze even a moderate amount of data manually. However, technological innovations over the past decade have automated previously manual tasks such as license plate reading, tracking individuals on video feeds, and other complex tasks. In these comparable situations, two separate but related technological innovations were required: inexpensive and automatic data collection hardware (e.g. via CCTV or video cameras), and automatic, accurate data processing software. We are in the process of developing a low-profile scanner which, when embedded in a pedestrian area, passively collects images of shoe treads (a prototype is shown in \autoref{fig:prototype-scanner}). In parallel, however, it is necessary to develop software which can process the data from a similarly oriented scanner to generate features which are used to make class characteristic comparisons. Through the combination of hardware and software, we hope to enable the collection of population-level footwear data which will facilitate quantitative assessment of class characteristic frequency to support examiner evaluation.

@@

Although footwear impressions are more commonly used in practice for footwear analysis than outsole images, our early prototypes of a footwear scanner suggest that cameras are more robust than impression-based systems (e.g.  EverOS) for use in outdoor settings. As a result, in this paper describing an initial automated software system for class characteristic recognition, we use outsole images (such as those found on online retail sites) rather than impression-based training data. While we recognize that images from the scanner are likely to show some degradation compared to images found from online retailers, the fundamental features of the underlying machine learning models are similar. In this paper, we present a proof-of-concept model designed to identify class characteristics probabilistically, using labels similar to those used by examiners. We discuss the strengths and limitations of both the classification scheme and the neural-network based model, with important implications for future attempts to develop automatic high-level class characteristic recognition systems.
